% smrdoc.tex V2.0, 13 May 2010
\documentclass[times]{smrauth}

% Added packages
\usepackage{url}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{color}
\usepackage{textcomp}
\usepackage{algorithm}
\usepackage{algpseudocode}
\usepackage{moreverb}
\usepackage[dvips,colorlinks,bookmarksopen,bookmarksnumbered,citecolor=red,urlcolor=red]{hyperref}


\floatname{algorithm}{Pseudo Code}
\def\sectionautorefname{section}
\def\subsectionautorefname{section}
\def\subsubsectionautorefname{section}
\def\figureautorefname{figure}

\definecolor{dark-gray}{rgb}{0.7,0.7,0.7}
\definecolor{gray}{rgb}{0.5,0.5,0.5}
\definecolor{light-gray}{rgb}{0.9,0.9,0.9}

\renewcommand{\algorithmiccomment}[1]{\color{gray}\emph{// #1}\color{black}}

\newcommand{\tvar}[1]{\emph{#1}}
\newcommand{\tdefine}[2]{\textbf{define} \tvar{#1} = #2}
\newcommand{\tquote}[1]{\textquotesingle\textquotesingle#1\textquotesingle\textquotesingle}
\newcommand{\tsingquote}[1]{\textquotesingle#1\textquotesingle}

\newcommand\BibTeX{{\rmfamily B\kern-.05em \textsc{i\kern-.025em b}\kern-.08em
T\kern-.1667em\lower.7ex\hbox{E}\kern-.125emX}}

\def\volumeyear{2010}

\lstdefinestyle{sqlStyle}{
  breaklines=true,                                     % line wrapping on
  language=SQL,
  basicstyle=\tiny,
  keywordstyle=\ttfamily\color{black}\bfseries,
  identifierstyle=\ttfamily\color{black},
  commentstyle=\color{gray},
  stringstyle=\ttfamily\color{dark-gray},
  showstringspaces=false,
  tabsize=2,
  extendedchars=true,
  upquote=true,
  backgroundcolor=\color{light-gray},
  captionpos=b,
  frame=single,
  frameround=tttt,
  fillcolor=\color{light-gray},
  float=h,
  morekeywords={AUTO_INCREMENT, bigint, REFERENCES, CHANGE, NOW, TIMESTAMPDIFF, SECOND}
}

\begin{document}

\runningheads{L. Prévost and O. Liechti}{}

\title{Database Evolution Framework}
%\title{A demonstration of the \LaTeXe\ class file for the
%\itshape{\journalnamelc}\footnotemark[2]}

\author{Laurent Prévost\affil{1,2}, Olivier Liechti\affil{1,2}}
\address{
\affilnum{1}University of Applied Sciences of Western Switzerland, 1400 Yverdon-les-Bains, Switzerland\break
\affilnum{2}Lotaris SA, Galilée 15, 1400 Yverdon-les-Bains, Switzerland}


%\corraddr{Journals Production Department, John Wiley \& Sons, Ltd,
%The Atrium, Southern Gate, Chichester, West Sussex, PO19~8SQ, UK.}

\begin{abstract}

In a service oriented architecture, dealing with the evolution of the underlying database is not a trivial problem. New functional requirements are often associated with modifications of the domain model, hence with modifications of the relational schema. These modifications have to be done in a way that does not impact existing services and applications. They also have to be done in a way that preserves consistency in the data store. Last but not least, they have to be done in a way that has minimal impact on the service availability. In this paper, we propose a framework that supports the development, maintenance and operational activities. The framework consists of processes, tools and best practices. It is the result of our practical experience with a large-scale commercial internet service, which is characterized both by a constant and rapid functional evolution and by strict operational constraints. 

\end{abstract}


\keywords{}

\maketitle

%\footnotetext[2]{Please ensure that you use the most up to date
%class file,
%available from the SMR Home Page at\\
%\href{http://www3.interscience.wiley.com/journal/117946198/grouphome/home.html}{\texttt{http://www3.interscience.wiley.com/journal/117946198/grouphome/home.html}}}

\renewcommand{\thefootnote}{\arabic{footnote}}

\input{chapters/intro.tex}
\input{chapters/related.tex}
\input{chapters/def.tex}
\input{chapters/risks.tex}
\input{chapters/casestudy.tex}
\input{chapters/futurework.tex}
\input{chapters/conclusion.tex}

\bibliographystyle{plain}
\bibliography{pa}
\end{document}
